---
title: "Project 1"
author: "Isabella Torres | Yuan Pei Yang"
date: "2026-04-05"
output: html_document
---

```{r datasets}
#Loading data sets and assigning names for use
confirmations <- read.csv("data/time_series_covid19_confirmed_global.csv", stringsAsFactors = FALSE)
deaths <- read.csv("data/time_series_covid19_deaths_global.csv", stringsAsFactors = FALSE)
```

# Objective 1: Covid's Origin

```{r obj1}

```

# Objective 2: Most Recent Area Exposed

```{r obj2}

```

# Objective 3: How Far the Virus Traveled

```{r obj3}

```

# Objective 4: Who is Most At Risk?

```{r obj4}
#Create vector of location information
locationData = c("Province.State","Country.Region","Lat","Long")

#Find last column of the data set to use most recent data for every render
lastCol = ncol(confirmations)

#Duplicate last column under a new name
confirmations$recentConf = confirmations[, lastCol]
deaths$recentDeath = deaths[, lastCol]

#Merge confirmations and deaths data sets
riskData = merge(x=confirmations, y=deaths,
                 by.x=c(locationData),
                 by.y=c(locationData))
#Take objective-specific columns
riskData = riskData[c(locationData,"recentConf","recentDeath")]

#Filter data to exclude cruise ships
riskData = riskData[!is.na(riskData$Lat) & riskData$Lat != 0, ]

#Calculate Risk
risk = 100 * (riskData$recentDeath / riskData$recentConf)
riskData$Risk.Score = risk #Add risk score to riskData 
riskData = riskData[!is.na(riskData$Risk.Score), ] #Remove NA values

#Global risk
globalRisk = 100 * (sum(deaths$X3.9.23, na.rm = TRUE) / sum(confirmations$X3.9.23, na.rm = TRUE))

#Find areas with the lowest risk
lowRisk = riskData[riskData$Risk.Score == min(risk, na.rm = TRUE), ] 
mostConf = max(lowRisk[, "recentConf"]) #Find most confirmations within lowRisk areas

#Area with the lowest risk:
lowRisk = lowRisk[lowRisk$recentConf == mostConf, ]

#Area with the highest risk:
highRisk = riskData[riskData$Risk.Score == max(risk, na.rm = TRUE), ] 

#Displaying results
cat("Global Risk:", globalRisk, "\n")
cat("Area with the Lowest Risk Score:", lowRisk[[1]], lowRisk[[2]],"-> Risk Score =", lowRisk[[7]], "\n")
cat("Area with the Highest Risk Score:", highRisk[[2]],"-> Risk Score =", highRisk[[7]])
```

**Which area of the world currently has the lowest risk score (if more than one, display the one with the most confirmations)? Which area of the world currently has the highest risk score (if more than one, display the one with the most confirmations)? How do risk scores in these areas compare to global risk score? Why might it be helpful to calculate metrics like risk scores for different areas of the world and what would their limitations be (what assumptions does risk score make and what important variables might be left out)?**

Write Up: Jiangsu, China has the lowest risk score, while North Korea has the highest risk score. The risk score of Jiangsu, China in comparison to the world only has a difference of about 1%, while North Korea's risk score differs by 599%. Risk scores can assume the healthcare industry of that area; with lower scores having better healthcare or better resources than areas with higher risk scores. We can use this information to know how to allocate resources based on need. Some limitations can be like North Korea, which showed the highest risk score. However, it is likely that they just did not submit the correct number of confirmations and deaths. In other words, data integrity is something data scientists need to consider in their analysis.

# Objective 5: Top 5 Countries

```{r obj5}
#Create vector of countries
countries = unique(confirmations$Country.Region)

#Load knitr package for kable
library("knitr")
```

### Top 5 Confirmations

```{r obj5_conf}
#Create empty vector to hold sum of confirmations per country
confSum = c()
confCol = names(confirmations)[ncol(confirmations)] #column that holds most recent count

for (i in countries) { #iterates through all countries
  #Subset of data frame using country specified in loop
  countrySubset = confirmations[confirmations$Country.Region == i, ]
  #Compute country's sum of confirmations
  countryConf = sum(countrySubset[[confCol]], na.rm = TRUE)
  #Adds country's sum to confSum vector
  confSum = c(confSum, countryConf)
}

#Create data frame of countries and their confirmation count
top5Conf = data.frame(Country = countries, Confirmations = confSum)
top5Conf = top5Conf[order(top5Conf$Confirmations, decreasing = TRUE), ] #Ordered by confirmations

#Take only top 5 countries with the most confirmations
top5Conf = head(top5Conf, n = 5)

#Return table, removes row indices
kable(top5Conf, row.names = FALSE)
```

### Top 5 Deaths

```{r obj5_death}
#Create empty vector to hold sum of deaths per country
deathSum = c()
deathCol = names(deaths)[ncol(deaths)] #column that holds most recent count

for (i in countries) { #iterates through all countries
  #Subset of data frame using country specified in loop
  countrySubset = deaths[deaths$Country.Region == i, ]
  #Compute country's sum of deaths
  countryDeath = sum(countrySubset[[deathCol]], na.rm = TRUE)
  #Adds country's sum to deathSum vector
  deathSum = c(deathSum, countryDeath)
}

#Create data frame of countries and their death count
top5Death = data.frame(Country = countries, Deaths = deathSum)
top5Death = top5Death[order(top5Death$Deaths, decreasing = TRUE), ] #Ordered by deaths

#Take only top 5 countries with the most deaths
top5Death = head(top5Death, n = 5)

#Return table, removes row indices
kable(top5Death, row.names = FALSE)
```
